MALTEM effectiveness analysis

### 2017-07-19 paper.Rmd
update

---
title: "Cost-effectiveness of a malaria elimination campaign"
author: "Laia Cirera, Joe Brew, Elisa Sicuri"
date: "`r Sys.Date()`"
output:
 tufte::tufte_html: default
 tufte::tufte_handout:
   citation_package: natbib
   latex_engine: xelatex
 tufte::tufte_book:
   citation_package: natbib
   latex_engine: xelatex
bibliography: bibliography.bib
link-citations: yes
---

@@

# Create variables 
bes <- 
  bes %>%
  mutate(mda=ifelse(date<"2015-11-08", "pre_mda", 
                    ifelse(date<"2016-02-04", "mda", "post_mda")),
         intervention=district=="MAGUDE",
         magude = ifelse(district == 'MAGUDE', 'Magude', 'Not Magude'))

# Create model to estimate effectiveness
mod <-lm(pk ~ mda*intervention + irs_coverage, data=bes, weights = population)
```

@@

counter_factual$p_cf <- predict(mod, counter_factual)
counter_factual$cases_cf <- counter_factual$p_cf*counter_factual$population 
counter_factual$cases_averted <- counter_factual$cases_cf - counter_factual$cases

@@

DT::datatable(total_intervention_cost)
```

>>>>

---
title: "Cost-effectiveness of a malaria elimination campaign"
author: "Laia Cirera, Joe Brew, Elisa Sicuri"
date: "`r March 31, 2017`"
output:
 tufte::tufte_html: default
 tufte::tufte_handout:
   citation_package: natbib
   latex_engine: xelatex
 tufte::tufte_book:
   citation_package: natbib
   latex_engine: xelatex
bibliography: bibliography.bib
link-citations: yes
---

@@

# Create variables 
bes <- 
  bes %>%
  mutate(mda=ifelse(date<"2015-08-01", "pre_maltem", 'maltem'),
         intervention=district=="MAGUDE",
         magude = ifelse(district == 'MAGUDE', 'Magude', 'Not Magude'))

# Create a time since 2010 variable
bes <- bes %>%
  mutate(time_since_2010 = as.numeric(date - as.Date('2010-01-01')))

# Make a season variable
bes <- bes %>%
  mutate(season = ifelse(month %in% 1:5, 'high', 'low'))

# Create model to estimate effectiveness
mod <-lm(pk ~ mda*intervention + 
           irs_coverage + 
           season +
           factor(month) +
           time_since_2010, 
         data=bes %>%
           filter(date >= '2012-01-01'), 
         weights = population)
```

@@

counter_factual$p_cf <- predict(mod, counter_factual)
counter_factual$cases_cf <-(counter_factual$p_cf/1000)*counter_factual$population 
counter_factual$cases_averted <- counter_factual$cases_cf - counter_factual$cases

@@

DT::datatable(total_intervention_cost)

# Plot of the counterfactual and real
real <- bes %>%
  filter(district == 'MAGUDE') %>%
  mutate(status = 'observed')
cf <- counter_factual %>%
  # filter(year == 2016) %>%
  filter(year < 2017) %>%
  mutate(status = 'counterfactual') %>%
  mutate(cases = cases_cf,
         p = p_cf)
plot_data <- bind_rows(real, cf)

x <- plot_data %>%
  filter(date >= '2015-08-01') %>%
  group_by(status, year) %>%
  summarise(total_cases = sum(cases))

days_of_observation <- as.numeric(as.Date('2016-12-31') - as.Date('2015-08-01'))

plot_data <- plot_data %>%
  group_by(year, week, status) %>%
  summarise(month = first(month),
            date = first(date),
            cases = sum(cases),
            population = sum(population)) %>%
  ungroup %>%
  mutate(p = cases / population) %>%
  mutate(pk = p * 1000)

library(cism)
plot_data <- plot_data %>%
 filter(date >= '2014-07-01')
g <- ggplot(data = plot_data ,
       aes(x = date,
           y = pk,
           group = status,
           color = status)) +
  geom_line(alpha = 0.6) +
  theme_cism() +
  labs(x = 'Date',
       y = 'Weekly incidence per 1000',
       title = 'Observed vs. counterfactual',
       subtitle = 'Preliminary effectivenesss model') +
  scale_color_manual(name = 'Status',
                     values = c('darkorange', 'darkgreen'))
g +
  geom_vline(xintercept = c(as.numeric(as.Date('2015-08-01')),
                            as.numeric(as.Date('2015-11-01'))),
             alpha = 0.5,
             lty = 2)

# By mes
pdm <- plot_data %>%
  mutate(date = as.Date(paste0(year, '-',
                               month, '-01'))) %>%
  group_by(date, status) %>%
  summarise(pk = mean(pk))
ggplot(data = pdm ,
       aes(x = date,
           y = pk,
           group = status,
           color = status)) +
  geom_line(alpha = 0.6) +
  theme_cism() +
  labs(x = 'Date',
       y = 'Monthly incidence per 1,000',
       title = 'Preliminary effectiveness model',
       subtitle = 'Magude district') +
  scale_color_manual(name = '',
                     values = c('darkorange', 'darkgreen')) +
    geom_vline(xintercept = c(as.numeric(as.Date('2015-08-01')),
                            as.numeric(as.Date('2015-11-09')),
                            as.numeric(as.Date('2016-12-02'))),
             alpha = 0.5,
             lty = 2)

# Get difference b/w 2016 counterfactual and real
x <- counter_factual %>%
  filter(year == 2016) %>%
  summarise(cf = sum(cases_cf, na.rm = TRUE),
            cr = sum(cases)) %>%
  mutate(cases_averted = cf - cr)

# write a csv for laia
write_csv(plot_data, '~/Desktop/cf_vs_observed.csv')
```

### 2017-07-19 reports/misc_analysis/misc_analysis.Rmd
update

fit <- lm(p ~ precipitation_lagged + 
            factor(month), data = x)
broom::tidy(fit)
```

@@

@@

>>>>

fit <- lm(p ~ precipitation_lagged + 
            factor(month), data = x)
DT::datatable(broom::tidy(fit))
```

@@

We calculate the distance from the centroid of each district to the South African border.

```{r}
# Interpolate for our locations
library(sp)
x <- cism::moz2
x <- x[x@data$NAME_1 %in% c('Maputo', 'Gaza'),]
coords <- data.frame(coordinates(x))
names(coords) <- c('x', 'y')
coords$district <- x@data$NAME_2

# Standardize names
coords$district <- toupper(coords$district)
coords$district[coords$district == 'CHÓKWÈ'] <- 'CHOKWE'
coords$district[coords$district == 'GUIJÁ'] <- 'GUIJA'
coords$district[coords$district == 'XAI-XAI'] <- 'XAI-XAI DISTRICT'
coords$district[coords$district == 'MANHIÇA'] <- 'MANHICA'
coords$district[coords$district == 'MATUTUÍNE'] <- 'MATUTUINE'
coords$district[coords$district == 'MANDLAKAZI'] <- 'MANJACAZE'

# Add a row for matola
matola <- data_frame(x = 32.46580,
                     y = -25.92391,
                     district = 'MATOLA')
coords <- bind_rows(coords, 
                    matola)
# Add a row for xai-xai city
xxc <- coords %>% filter(district == 'XAI-XAI DISTRICT') %>%
  mutate(district = 'XAI-XAI CITY')
coords <- bind_rows(coords,
                    xxc)
# Make spatial
coords <- coords %>%
  mutate(lng = x,
         lat = y)
coordinates(coords) <- ~x+y
proj4string(coords) <- proj4string(moz3)

# Get South Africa
sa <- cism::africa
sa <- sa[sa@data$COUNTRY == 'South Africa',]

# Get projected versions
sa_proj <- spTransform(sa,
                    CRS("+init=epsg:3347"))
coords_proj <- spTransform(coords,
                    CRS("+init=epsg:3347"))

# Get distance to south Africa
coords@data$distance_to_south_africa <- NA
for (i in 1:nrow(coords)){
  this_distance <- rgeos::gDistance(coords_proj[i,], sa_proj)
  coords@data$distance_to_south_africa[i] <- this_distance / 10000
}

afr <- cism::africa
afr <- afr[afr@data$COUNTRY %in% c('South Africa', 'Mozambique'),]
plot(afr, col = adjustcolor(c('blue', 'green'), alpha.f = 0.3))
points(coords, col = 'red',
       pch = '.',
       cex = 2)
```

The distances can be seen in the below table.

```{r}
DT::datatable(coords@data %>% dplyr::select(district, distance_to_south_africa) %>%
                mutate(distance_to_south_africa = round(distance_to_south_africa, digits = 2)) %>%
  rename(District = district,
         KM = distance_to_south_africa)) 
```

Visually, we assess the hypothesis that the decline over time is steeper for those districts bordering South Africa through a series of charts.

```{r}
x <- df %>%
  filter(disease == 'MALARIA') %>%
  left_join(coords@data) %>%
  group_by(district, year, distance_to_south_africa) %>%
  summarise(p = sum(cases) / sum(population) * 1000)

ggplot(data = x,
       aes(x = distance_to_south_africa,
           y = p)) +
  geom_point(alpha = 0.7) +
  geom_smooth(n = 5, span = 10, se = FALSE,
              color = 'darkred',
              alpha = 0.5,
              lty = 1) +
  facet_wrap(~year) +
  theme_maltem() +
  labs(x = 'Kilometers to South Africa',
       y = 'Weekly incidence per K',
       title = 'Protective effect of South African proximity',
       subtitle = 'One point per district per year')

As an alternative to the above, we can visualize the same data with the same aesthetics, but overlaying each year.

```{r}
library(RColorBrewer)
cols <- colorRampPalette(brewer.pal(n = 9, 'Spectral'))(length(unique(x$year)))
ggplot(data = x,
       aes(x = distance_to_south_africa,
           y = p,
           color = factor(year))) +
  geom_point(alpha = 0.7) +
  geom_smooth(n = 5, span = 10, se = FALSE,
              alpha = 0.5,
              lty = 1) +
  theme_maltem() +
  labs(x = 'Kilometers to South Africa',
       y = 'Weekly incidence per K',
       title = 'Protective effect of South African proximity',
       subtitle = 'One point per district per year') +
  scale_color_manual(name = 'Year',
                     values = cols)
```

Could the proximity to South Africa partly explain the pre-MALTEM decline in incidence in Magude? We'll examine this by modeling incidence as a linear function of district, distance, the interaction of year and distance, and month, in effect constructing a "counter-factual" for Magude that takes into account the secular decline in incidence associated with proximity to South Africa. For the model's purpose, we take into account only those 4 years from 2012-06-01 through 2016-06-01 (so as to not include the effect of MALTEM). Our model formula predicts incidence as a function of the interactions of season/year and distance/year, with the fixed effect of month.

```{r}
model_data <- 
  df %>%
  filter(disease == 'MALARIA') %>%
  left_join(coords@data) %>%
  mutate(season = ifelse(month %in% 1:4, 'rainy', 'not-rainy')) %>%
  mutate(month = factor(month)) %>%
  group_by(date, year, 
           season,
           month,
           district, distance_to_south_africa) %>%
  summarise(p = sum(cases) / sum(population) * 1000) %>%
  ungroup

fit <- lm(p ~ year*log(distance_to_south_africa)  + year + season*year + month, 
          data = model_data %>%
            filter(date <= '2016-06-01',
                   date >= '2012-06-01'))

DT::datatable(broom::tidy(fit))

model_data$predicted <- predict(fit, model_data)

ggplot() +
  geom_line(data = model_data %>%
         filter(district == 'MAGUDE'),
       aes(x = date,
           y = predicted),
       alpha = 0.8) +
  geom_line(data = model_data %>%
         filter(district == 'MAGUDE'),
       aes(x = date,
           y = p),
       color = 'red',
       alpha = 0.8) +
  labs(title = 'Weekly malaria incidence in Magude',
       subtitle = 'Observed (red) vs. predicted (black)',
       x = 'Date',
       y = 'Cases per 1,000') +
  theme_maltem()
```

# Effect of insecticide-treated bednets (ITN)

At first glance, ITN coverage appears associated with lower malaria incidence

```{r}
ggplot(data = df %>% filter(disease == 'MALARIA'),
       aes(x = itn_coverage,
           y = pk)) +
  geom_point(alpha = 0.3) +
  theme_maltem() +
  labs(x = 'ITN coverage',
       y = 'Weekly incidence per 1,000',
       title = 'ITN and malaria',
       subtitle = 'One point per district, week, age-group') +
  scale_x_log10()
```

# Effect of indoor residual spraying (IRS)

```{r}
x <- df %>%
  filter(disease == 'MALARIA') %>%
  group_by(date, district, irs_coverage_people, weeks_since_last_irs_campaign_end) %>%
  summarise(p = sum(cases) / sum(population) * 1000) %>%
  ungroup %>%
  mutate(irs_time_since = ifelse(weeks_since_last_irs_campaign_end <= 10,
                                 '00-10',
                                 ifelse(weeks_since_last_irs_campaign_end <= 50,
                                        '11-50',
                                        ifelse(weeks_since_last_irs_campaign_end <= 99,
                                               '51-99',
                                               ifelse(weeks_since_last_irs_campaign_end > 100,
                                                      '99+',
                                                      NA))))) %>%
  filter(!is.na(weeks_since_last_irs_campaign_end)) %>%
  mutate(irs_coverage = ifelse(irs_coverage_people < 50,
                               '00-50%',
                               ifelse(irs_coverage_people <= 70,
                                      '51-70%',
                                      ifelse(irs_coverage_people <= 90,
                                             '71-90%',
                                             ifelse(irs_coverage_people > 90,
                                                    '91+%',
                                                    '00%')))))

ggplot(data = x,
       aes(x = weeks_since_last_irs_campaign_end,
           y = p)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~irs_coverage) +
  labs(x = 'Weeks since last IRS campaign end',
       y = 'Incidence per 1,000',
       title = 'Effect of IRS',
       subtitle = 'Faceted by coverage percentage') +
  theme_maltem()

# Incorporating more variables

Here's what happens when we model cases using ITN, IRS, time since IRS campaign, seasonality, etc.

```{r}
model_data <- 
  df %>%
  filter(disease == 'MALARIA') %>%
  left_join(coords@data) %>%
  mutate(season = ifelse(month %in% 1:4, 'rainy', 'not-rainy')) %>%
  mutate(month = factor(month)) %>%
  group_by(date, year, 
           season,
           month,
           irs_coverage_people,
           weeks_since_last_irs_campaign_end,
           itn_coverage,
           district, distance_to_south_africa) %>%
  summarise(p = sum(cases) / sum(population) * 1000) %>%
  ungroup

fit <- lm(p ~ year*log(distance_to_south_africa)  + year + season*year  + month +  irs_coverage_people*weeks_since_last_irs_campaign_end + itn_coverage + district, 
          data = model_data %>%
            filter(date <= '2016-06-01',
                   date >= '2012-06-01'))

x <- model_data %>% filter(district == 'MAGUDE')
x$predicted <- predict(fit, x)

ggplot() +
  geom_line(data = x,
       aes(x = date,
           y = predicted),
       alpha = 0.8) +
  geom_line(data = x,
       aes(x = date,
           y = p),
       color = 'red',
       alpha = 0.8) +
  labs(title = 'Weekly malaria incidence in Magude',
       subtitle = 'Observed (red) vs. predicted (black)',
       x = 'Date',
       y = 'Cases per 1,000') +
  theme_maltem()

@@

2. Given that proximity to South Africa seems to have an interactive effect with the decline in malaria incidence, should we include this in our model? If so, how?

3. Are we currently excluding Manhiça from our synthetic controls? If so, why?

4. What are we doing with IRS / ITN?

5. How are we treating secular trends (ie, near linear decline in springtime malaria in Magude for the 4 years from 2012-2015)?
